---
title: "Homework 6 "
author: "Wenwen Li"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(broom)
library(purrr)
library(corrplot)
library(gridExtra)
library(caret)
library(modelr)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```
### Problem 1

In the data cleaning code below we create a `city_state` variable, change `victim_age` to numeric, modifiy victim_race to have categories white and non-white, with white as the reference category, and create a `resolution` variable indicating whether the homicide is solved. Lastly, we filtered out the following cities: Tulsa, AL; Dallas, TX; Phoenix, AZ; and Kansas City, MO; and we retained only the variables `city_state`, `resolution`, `victim_age`, `victim_sex`, and `victim_race`.

```{r q1_data_cleaning}
homicide_df = 
  read_csv("data/homicide-data.csv", na = c("", "NA", "Unknown")) |> 
  mutate(
    city_state = str_c(city, state, sep = ", "),
    victim_age = as.numeric(victim_age),
    resolution = case_when(
      disposition == "Closed without arrest" ~ 0,
      disposition == "Open/No arrest"        ~ 0,
      disposition == "Closed by arrest"      ~ 1)
  ) |> 
  filter(victim_race %in% c("White", "Black")) |> 
  filter(!(city_state %in% c("Tulsa, AL", "Dallas, TX", "Phoenix, AZ", "Kansas City, MO"))) |> 
  select(city_state, resolution, victim_age, victim_sex, victim_race)
```

Next we fit a logistic regression model using only data from Baltimore, MD. We model `resolved` as the outcome and `victim_age`, `victim_sex`, and `victim_race` as predictors. We save the output as `baltimore_glm` so that we can apply `broom::tidy` to this object and obtain the estimate and confidence interval of the adjusted odds ratio for solving homicides comparing non-white victims to white victims.

```{r q1_glm_baltimore}
baltimore_glm = 
  filter(homicide_df, city_state == "Baltimore, MD") |> 
  glm(resolution ~ victim_age + victim_sex + victim_race, family = binomial(), data = _)

baltimore_glm |> 
  broom::tidy() |> 
  mutate(
    OR = exp(estimate), 
    OR_CI_upper = exp(estimate + 1.96 * std.error),
    OR_CI_lower = exp(estimate - 1.96 * std.error)) |> 
  filter(term == "victim_sexMale") |> 
  select(OR, OR_CI_lower, OR_CI_upper) |>
  knitr::kable(digits = 3)
```

Below, by incorporating `nest()`, `map()`, and `unnest()` into the preceding Baltimore-specific code, we fit a model for each of the cities, and extract the adjusted odds ratio (and CI) for solving homicides comparing non-white victims to white victims. We show the first 5 rows of the resulting dataframe of model results.

```{r q1_glm_all_cities}
model_results = 
  homicide_df |> 
  nest(data = -city_state) |> 
  mutate(
    models = map(data, \(df) glm(resolution ~ victim_age + victim_sex + victim_race, 
                             family = binomial(), data = df)),
    tidy_models = map(models, broom::tidy)) |> 
  select(-models, -data) |> 
  unnest(cols = tidy_models) |> 
  mutate(
    OR = exp(estimate), 
    OR_CI_upper = exp(estimate + 1.96 * std.error),
    OR_CI_lower = exp(estimate - 1.96 * std.error)) |> 
  filter(term == "victim_sexMale") |> 
  select(city_state, OR, OR_CI_lower, OR_CI_upper)

model_results |>
  slice(1:5) |> 
  knitr::kable(digits = 3)
```

Below we generate a plot of the estimated ORs and CIs for each city, ordered by magnitude of the OR from smallest to largest. From this plot we see that most cities have odds ratios that are smaller than 1, suggesting that crimes with male victims have smaller odds of resolution compared to crimes with female victims after adjusting for victim age and race. This disparity is strongest in New yrok. In roughly half of these cities, confidence intervals are narrow and do not contain 1, suggesting a significant difference in resolution rates by sex after adjustment for victim age and race. 

```{r q1_plot}
model_results |> 
  mutate(city_state = fct_reorder(city_state, OR)) |> 
  ggplot(aes(x = city_state, y = OR)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = OR_CI_lower, ymax = OR_CI_upper)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
### Problem 2
## Accessing the data
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2022-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

```{r}
#function to calculate R-squared and log(beta1 * beta2)
calculate_metrics = function(data) {
  model = lm(tmax ~ tmin + prcp, data = data)
  tidy_results = broom::tidy(model)
  r_squared = broom::glance(model)$r.squared
  #taking the logarithm after converting to absolute values
  beta_product = log(tidy_results$estimate[2]*
                        tidy_results$estimate[3])
  metrics_df = data.frame(R_squared = r_squared,
                           Beta_product = beta_product)
  return(metrics_df)
}

calculate_metrics(weather_df)

```

```{r}
#function to extract R-squared and compute log(beta1*beta2) for each bootstrap replicate
calculate_metrics = function(data) {
  data= data.frame(data)
  model = lm(tmax ~ tmin + prcp, data = data)
  tidy_results = broom::tidy(model)
  r_squared = broom::glance(model)$r.squared
  #taking the logarithm of beta1*beta2
  beta_product = log(tidy_results$estimate[2] * tidy_results$estimate[3])
  #return a data frame for the current replicate
  data.frame(R_squared = r_squared, Beta_product = beta_product)
}

#5000 bootstraps using modelr's bootstrap function
boot = modelr::bootstrap(weather_df, 5000)

#use the custom tidying function
models = map(boot$strap, calculate_metrics)

#combine the results into a data frame
bootstrap_results = bind_rows(models, .id = "Replicate")

#view the results
head(bootstrap_results)
```

```{r}
#visualize bootstrap distribution of R-squared
ggplot(bootstrap_results, aes(x = R_squared)) +
  geom_histogram(binwidth = 0.005, alpha = 0.7) +
  labs(title = "Bootstrapped distribution of R-squared",
       x = "R-squared")
```

The bootstrap distribution of the R-squared metric from the 5000 bootstrap samples assumes a slight left-skewed distribution, from the shape of the histogram.  


```{r, warning=FALSE}
#visualize bootstrap distribution of log(beta1 * beta2)
ggplot(bootstrap_results, aes(x = Beta_product)) +
  geom_histogram(binwidth = 0.1, alpha = 0.7) +
  labs(title = "Bootstrapped distribution of log(beta1 * beta2)",
       x = "log(beta1 * beta2)")

```

The bootstrap distribution of the log(beta1 * beta2) value from the 5000 bootstrap samples indicates a heavy left-skewed distribution.   


```{r}
#compute 95% confidence intervals for R-squared and log(beta1*beta2)
confidence_intervals = summarize(bootstrap_results,
  CIlower_R2 = quantile(R_squared, 0.025),
  CIupper_R2 = quantile(R_squared, 0.975),
  CIlower_b1xb2 = quantile(Beta_product, 0.025, na.rm=TRUE),
  CIupper_b1xb2 = quantile(Beta_product, 0.975, na.rm=TRUE)) 

confidence_intervals

```

# Problem 3
## Data Importation and preparation

```{r}
#import birthweight data
birthweight_data = read.csv("data/birthweight.csv")
#display the structure of the dataset
str(birthweight_data)
```

```{r}
#check for missing data
sum(is.na(birthweight_data))
```
